

In this assessment we will define error controlling procedures for experimental data. We will make list of genes based on q-values. 
We will also assess your understanding of false positives rates and false negative rates by asking you to create a Monte Carlo simulation.

You will need to install the following libraries:
        
```{r}
setwd("/mnt/hd2tb/Documents/moocs/dataanaly-lifesci/03-advinference")        
#library(devtools)
library(rafalib)

#install_github("genomicsclass/GSE5859Subset")
#install_bioc("genefilter")
#install_bioc("qvalue")
library(qvalue)
```

FDR Exercises #1 (1 point possible)

Load the gene expression data
```{r}


library(GSE5859Subset)
data(GSE5859Subset)

```
We are interested in comparing gene expression between the two groups defined in the sampleInfo table.

Compute a p-value for each gene using the function rowttests from the genefilter package in Bioconductor.

```{r}


library(genefilter)
#?rowttests
set.seed(1)
n <- nrow(geneExpression)
nullHypothesis <- c( rep(TRUE,n))
##Define groups to be used with rowttests
g <- factor(sampleInfo$group )
B <- 1 ##number of simulations
calls_vec <-0
system.time(
        VandS <- replicate(B,{
                ##matrix with control data (rows are tests, columns are mice)

                dat <- geneExpression
                
                calls_vec <<- rowttests(dat,g)$p.value < alpha
                
                #c(sum(nullHypothesis & calls),sum(!nullHypothesis & calls))
                cat("V = ",sum(nullHypothesis & calls_vec),
                    "S = ",sum(!nullHypothesis & calls_vec), "\n")
                
        })
)
```

How many genes have p-values smaller than 0.05?
```{r}

table(calls_vec, nullHypothesis)

```


FDR Exercises #2 (1 point possible)

Apply the Bonferroni correction to the p-values obtained in question #1 to achieve a FWER of 0.05. How many genes are called significant under this procedure?

```{r}


bonf <- alpha/n
table(rowttests(dat,g)$p.value < bonf)


```
Note that we went from over a thousand to just 10. One practical question we should ask is "are we being too conservative?". Note that we have not really assessed false negatives at all yet. 


FDR Exercises #3 


Note that the FDR is a property of a list of features, not each specific feature. The q-value relates FDR to an individual feature. To define the q-value we order features we tested by p-value then compute the FDRs for a list with the most significant, the two most significant, the three most significant, etc... The FDR of the list with the, say, m most significant tests is defined as the q-value of the m-th most significant feature. In other words, the q-value of a feature, is the FDR of the biggest list that includes that gene.

In R, we can compute the q-value using the p.adjust function with the FDR option. Read the help file for p.adjust and then, for our gene expression dataset, compute how many genes achieve an FDR < 0.05


FDR Exercises #4 

Now use the qvalue function, in the Bioconductor qvalue package, to estimate q-values using the procedure described by Storey.

Using this estimate how many genes have q-values below 0.05?


FDR Exercises #5

Read the help file for qvalue and report the estimated proportion of genes for which the null hypothesis is true π0=m0/m


FDR Exercises #6 

Note that we have the number of genes passing the q-value <0.05 threshold is larger with the qvalue function than the p.adjust difference.

Why is this the case? Make a plot of the ratio of these two estimates to help answer the question.
One of the two procedures is flawed.
The two functions are estimating different things.
The qvalue function estimates the proportion of genes for which the null hypothesis is true and provides a less conservative estimate
The qvalue function estimates the proportion of genes for which the null hypothesis is true and provides a more conservative estimate


FDR Exercises #7


Note that this is an advanced question and that you can ask questions in the discussion forum.

Create a Monte Carlo Simulation in which you simulate measurements 
from 8,793 genes for 24 samples: 12 cases and 12 controls.
```{r}


n <- 24
m <- 8793
mat <- matrix(rnorm(n*m),m,n)

```
Now for 500 genes, there is a difference of 2 between cases and controls:

```{r}       
delta <- 2
positives <- 500
mat[1:positives,1:(n/2)] <- mat[1:positives,1:(n/2)]+delta


```

So the null hypothesis is true for 8793-500 genes. Using the notation from the videos 

```{r}


m=8793
m0=8293
m1=500
set.seed(1)

```
Set the seed at 1, set.seed(1) and run this experiment 1,000 times with a Monte Carlo simulation. For each instance compute p-values using a t-test (using rowttests in the genefilter package) and create three lists of genes using:
        
        Bonferroni correction to achieve an FWER of 0.05,
p-adjust estimates of FDR to achieve an FDR of 0.05, and
qvalue estimates of FDR to to achieve an FDR of 0.05.

For each of these three lists compute the number of false positives in the list and the number of false negatives: genes not in the list that should have been because the null hypothesis is not true (we added 2 to the controls).

What is the false positive rate (false positives divided by m0) if we use Bonferroni?

FDR Exercises #8 (1 point possible)

From the same Monte Carlo simulation as in the question above, what is the false negative rate if we use Bonferroni?


FDR Exercises #9 (1 point possible)

From the same Monte Carlo simulation as in question #7, what is the false positive rate if we use q-values from p.adjust?

FDR Exercises #10 (1 point possible)

From the same Monte Carlo simulation as in question #7, what is the false negative rate if we use q-values from p.adjust?

FDR Exercises #11 (1 point possible)

From the same Monte Carlo simulation as in question #7, what is the false positive rate if we use q-values from qvalue function?

FDR Exercises #12 (1 point possible)

From the same Monte Carlo simulation as in question #7, what is the false negative rate if we use q-values from qvalue function?